### 2020-07-12 paper.tex
add example of a density

%\newcommand{\disc}{\mathrm{disc}}
%\newcommand{\cont}{\mathrm{cont}}
%\newcommand{\tr}{\mathtt{tr}}
%\newcommand{\model}{\mathcal{P}}
%\newcommand{\proposal}{\mathcal{Q}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\newcommand{\true}{{\mathtt{T}}}
\newcommand{\false}{{\mathtt{F}}}

@@

% measure version
\subsection{Handling Continuous Distributions}
For programs that can sample from continuous probability distributions, the formalism needs to be extended.
Let $\mathcal{A}$ denote a set of addresses.
For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in A} V_a\}\}$ denote the set of all finite traces (i.e. dictionaries) from addresses in $\mathcal{A}$ to their values.
For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.

A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on traces.
Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper.
We refer to the reader to~\citet{popl} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.

This formalism is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
For example, consider the following probabilistic program:

TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and show the reference measure on traces and its density.
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function foo(x1, x2, x3)
  a = ({:a} ~ bernoulli(x1))
  b = ({:b} ~ bernoulli(x2))
  c = ({:c} ~ bernoulli(x3))
  {:d} ~ bernoulli((a && b && c) ? 0.9 : 0.1)
  return a && b && c && d
end
\end{lstlisting}}
\end{tabular}
\end{center}

TODO: somewhere, talk about hierarchical addresses
%\paragraph{Hierarchical addresses}

>>>>

%\newcommand{\disc}{\mathrm{disc}}
%\newcommand{\cont}{\mathrm{cont}}
%\newcommand{\tr}{\mathtt{tr}}
%\newcommand{\model}{\mathcal{P}}
%\newcommand{\proposal}{\mathcal{Q}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\newcommand{\true}{{\mathtt{T}}}
\newcommand{\false}{{\mathtt{F}}}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\naturals}{\mathbb{N}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\normal}{\mathcal{N}}
\newcommand{\unif}{\mathcal{U}}

@@

\subsection{Handling Continuous Distributions}
For programs that can sample from continuous probability distributions, the formalism needs to be generalized.
Instead of a probability mass function on traces, a probabilistic program defines a probability density function on traces with respect to a certain reference measure.

Let $\mathcal{A}$ denote a countable set of all possible addresses.
For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in A} V_a\}\}$ denote the set of all finite traces (i.e. dictionaries) from addresses in $\mathcal{A}$ to their values.
For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.

A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on traces.
Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper, and we refer to the reader to~\citet{popl} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.

The measure-theoretic formalism above is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
For example, consider the following probabilistic program, which encodes a generative model for data generated from a polynomial with unknown degree and unknown coefficients:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function polynomials(x_coords)
    degree ~ |geometric(0.5)|
    noise ~ |inv_gamma(1, 1)|
    coeffs = [ ({("c", i)} ~ |normal(0, 2)|)
               for i in 0:degree ]
    for i=1:length(x_coords)
        x = x_coords[i]
        mu = coeffs' * x.^(0:length(coeffs)-1)
        {("y", i)} ~ |normal(mu, noise + 0.1)|
    end
end
\end{lstlisting}}
\end{tabular}
\end{center}
The addresses used by this generative function are:
\[
\mathcal{A} := \begin{array}{c}
\{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
\{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
\{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
\end{array}
\]
(Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
Address $\mathtt{:}\mathtt{:noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and all of these use the Lebesgue measure. % TODO double-check this
The density with respect to the reference measure $\mu$ is:
\begin{align*}
p(x) =
\begin{array}{l}
p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
\prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
\prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
\end{array}
\end{align*}
where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$.
